---
title: "Feature 1 — Indigenous community eligibility summary"
subtitle: "FEMA disaster declarations, calendar years 2014–2026; program eligibility windows vary by program"
format:
  html:
    toc: true
    df-print: kable
execute:
  warning: true
  message: true
---

Summarizes each Indigenous community's history of presidentially-declared disasters. 

Set the `rerun` flag below to `TRUE` to re-execute the full pipeline;
otherwise this document reads the most recent cached outputs from
`data/fema/`.

```{r setup}
library(tidyverse)
library(arrow)
library(sf)
library(here)
library(janitor)
```

```{r}
params = list(rerun = TRUE)
```

```{r run-pipeline}
find_dated_files <- function(stem, ext = "parquet") {
  list.files(
    here("data", "fema"),
    pattern = str_c("^feature_1_", stem, "_\\d{4}_\\d{2}_\\d{2}\\.", ext, "$"),
    full.names = TRUE)
}

needs_rerun <- isTRUE(params$rerun) ||
  length(find_dated_files("eligibility")) == 0 ||
  length(find_dated_files("eligibility_detail")) == 0 ||
  length(find_dated_files("universe_sf", ext = "rds")) == 0 ||
  length(find_dated_files("counties_sf", ext = "rds")) == 0
if (needs_rerun) {
  source(here::here("scripts", "feature_1", "00_run_feature_1.R"))
  run_feature_1(refresh_fema = TRUE)
}
```

Read the most recent data files from the local cache (generated/updated in the chunk above.)

```{r load-outputs}
latest_for <- function(stem, ext = "parquet") {
  paths <- find_dated_files(stem, ext = ext)
  tibble(path = paths) %>%
    mutate(snapshot_date = path %>% str_extract("\\d{4}_\\d{2}_\\d{2}") %>%
             str_replace_all("_", "-") %>% ymd()) %>%
    slice_max(snapshot_date, n = 1) %>%
    pull(path)
}

tier_levels <- c("direct_named_match", "direct_spatial_county_match",
                 "statewide_state_match", "no_match")

classified <- read_parquet(latest_for("classified_declarations"))
tribal_declarations <- read_parquet(latest_for("tribal_declarations"))
county_declarations <- read_parquet(latest_for("county_declarations"))
statewide_declarations <- read_parquet(latest_for("statewide_declarations"))
universe <- read_parquet(latest_for("universe"))
results <- read_parquet(latest_for("eligibility")) %>%
  mutate(tier = factor(tier, levels = tier_levels, ordered = TRUE))

results_detail <- arrow::read_parquet(latest_for("eligibility_detail")) %>%
  collect() %>%
  mutate(tier = factor(tier, levels = tier_levels, ordered = TRUE))
universe_sf <- readRDS(latest_for("universe_sf", ext = "rds"))
counties_sf <- readRDS(latest_for("counties_sf", ext = "rds"))
```

Loaded intermediate and final outputs from `data/fema/`.

# Feature 1 - Dataset: Indigenous Community Search

One row per Indigenous community. The EDA-declaration columns are
emitted as three slots (`_1`, `_2`, `_3`) populated with the three most
recent qualifying declarations.

```{r feature1-search-spec}
final_columns_search = c(
  "community_id", ## a unique identifier for the community
  "community_name", ## name of the community
  "community_name_short", ## shortened community name (<=9 words where possible) for narrow/mobile displays
  "community_state", ## primary state in which the community is located (postal abbrev)
  "community_edr", ## EDA Economic Development Representative for the state (semicolon-joined when a state has multiple)
  "community_edr_email", ## email address of the EDR (semicolon-joined when a state has multiple; order matches `community_edr`)
  "eda_factor_eligibility", ## "yes" if any declaration dated 2023-2024 reaches the community by any path; "no" otherwise (re-derived within the EDA window, not read off the 2014-2026 tier)
  ## Three most-recent declarations in the 2023-2024 EDA window, slot 1 = most recent:
  "eda_declaration_name_1",
  "eda_declaration_name_2",
  "eda_declaration_name_3",
  "eda_declaration_year_1",
  "eda_declaration_year_2",
  "eda_declaration_year_3",
  ## FEMA disaster number of the declaration in each slot:
  "eda_declaration_disaster_number_1",
  "eda_declaration_disaster_number_2",
  "eda_declaration_disaster_number_3",
  ## Up to three counties named in slot N's declaration, formatted "County, ST" and joined with "; ":
  "eda_spatial_intersection_counties_1",
  "eda_spatial_intersection_counties_2",
  "eda_spatial_intersection_counties_3",
  "fema_hmgp_binary",
  "fema_hmgp_lead_applicant_declaration_binary",
  "fema_hmgp_lead_applicant_state_year",
  "sba_binary" ## "yes" if a qualifying-tier declaration's incident end date falls in 2025 or 2026
  )
```

```{r feature1-search-build}
source(here("scripts", "feature_1", "05_build_feature1_search.R"))
source(here("scripts", "feature_1", "07_get_feature1_search_codebook.R"))

feature1_search <- build_feature1_search(
  eligibility_detail = results_detail,
  classified_declarations = classified,
  counties_sf = counties_sf)

## Communities with no state cannot be assigned an EDA regional director, so
## they are excluded from the search dataset. This should never happen: warn
## loudly rather than letting a community disappear from the published file.
feature1_search_missing_state = feature1_search %>%
  filter(is.na(community_state))

if (nrow(feature1_search_missing_state) > 0) {
  warning(
    "Excluding ", nrow(feature1_search_missing_state),
    " community/communities from the search dataset for a missing state: ",
    str_c(feature1_search_missing_state$community_name, collapse = "; "),
    call. = FALSE)
}

feature1_search_data = feature1_search %>%
  mutate(across(matches("binary"), ~ case_when(.x == "yes" ~ 1, .x == "no" ~ 0, TRUE ~ 99))) %>%
  filter(!is.na(community_state))

feature1_search_codebook = get_feature1_search_codebook()

stopifnot(setequal(colnames(feature1_search_data), feature1_search_codebook$variable))
stopifnot(setequal(colnames(feature1_search_data), final_columns_search))

colnames(feature1_search_data)[!colnames(feature1_search_data) %in% final_columns_search]
final_columns_search[!final_columns_search %in% colnames(feature1_search_data)]
```

# Feature 1 - Dataset: Indigenous Community Disasters (Data Visualization)

```{r}
final_columns_data_visualization = c(
  "community_id",
  "community_name",
  "community_name_short",
  "disaster_id",
  "disaster_title",
  "disaster_year_declared",
  "disaster_type",
  "disaster_designated_area",
  "disaster_lead_applicant")
```

```{r feature1-data-visualization-build}
source(here("scripts", "feature_1", "06_build_feature1_data_visualization.R"))
source(here("scripts", "feature_1", "08_get_feature1_data_visualization_codebook.R"))

## Display formatting ("State of <name>", "Statewide declaration",
## " (Tribal area)" tags) is applied inside
## build_feature1_data_visualization(), so the shipped CSV and the codebook
## describe the same values.
feature1_data_visualization <- build_feature1_data_visualization(
    eligibility_detail = results_detail,
    classified_declarations = classified,
    counties_sf = counties_sf)

feature1_data_visualization_codebook <- get_feature1_data_visualization_codebook()

stopifnot(setequal(
  colnames(feature1_data_visualization),
  feature1_data_visualization_codebook$variable))
```

## Validation and write

```{r feature1-validate-and-write}
#' Stop when a column that must always be populated contains missing values
#'
#' @param data Tibble. The deliverable being checked.
#' @param required_columns Character vector. Columns that must be complete.
#' @param dataset_label Character. Name of the deliverable, used in the error.
#' @return `invisible(NULL)`. Called for the error it raises.
stop_on_missing_required_values = function(data,
                                           required_columns,
                                           dataset_label) {
  missing_counts = data %>%
    summarize(across(all_of(required_columns), ~ sum(is.na(.x)))) %>%
    pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
    filter(n_missing > 0)

  if (nrow(missing_counts) > 0) {
    stop(
      "Missing values in required column(s) of the ", dataset_label, ": ",
      str_c(missing_counts$column, " (", missing_counts$n_missing, ")",
            collapse = "; "),
      call. = FALSE)
  }
  invisible(NULL)
}

## Every community in the universe reaches the search deliverable. The search
## data drops any community with no state, so a community losing its state
## upstream shows up here as a hard failure rather than a quiet omission.
communities_missing_from_search_data = setdiff(
  results_detail$community_id, feature1_search_data$community_id)
communities_added_to_search_data = setdiff(
  feature1_search_data$community_id, results_detail$community_id)

if (length(communities_missing_from_search_data) > 0) {
  stop(
    "Communities in the resolved universe are missing from the search data: ",
    str_c(communities_missing_from_search_data, collapse = ", "),
    call. = FALSE)
}
if (length(communities_added_to_search_data) > 0) {
  stop(
    "Communities in the search data are absent from the resolved universe: ",
    str_c(communities_added_to_search_data, collapse = ", "),
    call. = FALSE)
}

## Every community appearing in the visualization data also appears in the
## search data. This is containment rather than set equality, because a
## community with no declarations in the 2014-2026 window has no visualization
## rows at all while still belonging in the search data. It catches a community
## being dropped from one deliverable but not the other — the failure that
## removed "Pulikla Tribe of Yurok People" from the search data while leaving
## its five visualization rows in place.
communities_missing_from_search = setdiff(
  feature1_data_visualization$community_id,
  feature1_search_data$community_id)

if (length(communities_missing_from_search) > 0) {
  stop(
    "Communities present in the visualization data are missing from the ",
    "search data: ",
    str_c(communities_missing_from_search, collapse = ", "),
    call. = FALSE)
}

## The search deliverable is one row per community, so a duplicated key means
## a join upstream fanned out and inflated the published file.
duplicated_search_community_ids = feature1_search_data$community_id[
  duplicated(feature1_search_data$community_id)]

if (length(duplicated_search_community_ids) > 0) {
  stop(
    "Duplicated community_id in the search data: ",
    str_c(unique(duplicated_search_community_ids), collapse = ", "),
    call. = FALSE)
}

## Columns that carry no meaning when empty: identifiers, display names, and
## the state that drives the EDR assignment. The declaration-slot columns are
## deliberately excluded — they are NA whenever a community has fewer than
## three qualifying declarations.
stop_on_missing_required_values(
  feature1_search_data,
  c("community_id", "community_name", "community_name_short",
    "community_state"),
  "search data")
stop_on_missing_required_values(
  feature1_data_visualization,
  c("community_id", "disaster_id", "disaster_year_declared"),
  "visualization data")

## All four deliverables are written together, only once every check above
## has passed.
snapshot_date = Sys.Date() %>% str_replace_all("-", "_")
interim_dir = here("data", "interim")
if (!dir.exists(interim_dir)) dir.create(interim_dir, recursive = TRUE)
interim_path = function(stem) {
  file.path(interim_dir, str_c(stem, "_", snapshot_date, ".csv"))
}
write_csv(feature1_search_data, interim_path("feature1_search_data"))
write_csv(feature1_search_codebook, interim_path("feature1_search_codebook"))
write_csv(feature1_data_visualization,
          interim_path("feature1_data_visualization"))
write_csv(feature1_data_visualization_codebook,
          interim_path("feature1_data_visualization_codebook"))
```